Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-28 manuscript/paper.tex
fixed labels for main figure 2 fixed figure caption fixed typo in sweep simulation
Closes #6

\section{Introduction} 
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a recombining
species, and are currently the subject of intense research 
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
gunnarsson2024scalable,deng2024robust} that practical application has 
become possible. Varied applications have been recently proposed,
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and 
the spatial 
location of genetic
ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
more powerful approaches to quantifying genetic 
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association 
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree}.
While these developements are exciting, the performance 
of these new methods depend critically on the accuracy of the inferred ARGs,
and although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data is 
understudied. In particular, there is a critial lack of software 
infrastructure to support evaluation and quality control of 
inferred ARGs.

Visualisation is fundamentally important to data analysis.
Many specialised tools exist to aid the visual analysis and 
quality control
of assembly~\citep{wick2015bandage,challis2020blobtoolkit},
read mapping~\citep{robinson2011integrative},
and variant
calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
for example.
At every stage of a bioinformatics pipeline, it is important to visualise 
results to avoid artefacts and aid understanding of the data.
Genome browsers such as IGV~\citep{robinson2011integrative} and the 
UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
data modalities, and are vital infrastructure for the field.
% . In the past few decades,
% genome browsers have become synonymous with sequence visualisation, by
% providing user-friendly, interactive interfaces to navigate genomic sequences
% and annotations. From early standalone tools to manage \textit{C. elegans} data
% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
% with the changing landscape of genomics. Modern genome browsers can load
% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
% The data security and scalability offered by this architecture ensures that
% genome browsers continue to be relevant in the biobank era.

@@

\subsection{Data model} \label{subsec:Data_Model} 
\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
ARGS~\citep{wong2024general}. This efficient ARG encoding 
is implemented by the 
\texttt{tskit} library~\citep{ralph2020efficiently}
and supported by most modern ARG
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
haller2019tree,baumdicker2022efficient,
adrion2020community,lauterbur2023expanding,korfmann2023weak,
tsambos2023link,tagami2024tstrait},
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
In \texttt{tskit}, ARGs are encoded as a
collection of tables, storing information about the 
nodes and edges that describe the graph topology and the sites 
and mutations encoding the sequence variation (Fig.~\ref{fig:Figure_1}).
% For example, the Node table describes sampled or ancestral sequences
% and records their time of birth, and the individual and population they belong
% to. The Edge table describes the relationship between pairs of nodes, and
% records the node IDs and the genomic locations over which the edge spans. The
% Mutation table records mutation events in the history of the sample, including
% the change of allelic state, time of event and node in which the event
% occurred. 
% The Node and Edge tables describe the genealogical relationships
% between the sampled and ancestral sequences, 
% while the Mutation table describes
% sequence variation contained in the samples. 
% (see \hyperref[subsec:User_Interface]{User Interface}).

@@

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges 
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale 
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider 
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interatively navigates the ARG.
This approach allows us to summarise very large ARGs interactively;
Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of 
\texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
ARG with around 2.7 million nodes and edges.

@@

\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
\centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
\caption{Screenshot of the Edges tab in \texttt{tsbrowse}.
Visualisation of the edges in a simulated ARG with a strong selective
sweep in the middle of the genome (see Supplement for details).
Edges are shown in the main browser pane on the left, with additional
histograms summarising the edges on the right.
The effect of the sweep can be seen by the lack of edges crossing the 
centre of the simulated chromosome. Other tabs are accessed by the 
buttons on the top bar.
\label{fig:Figure_2}}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with pages to
describe various aspects of the ARG. 
Figure~\ref{fig:Figure_2}
demonstrates the user interface using the Edges page as an example. 
The plot on the left is an overview of the 33,929 edges 
(parent-child relationships between nodes along the genome)
in a simulated ARG with a strong selective sweep (see
Sec.~\ref{subsec:sweep_simulation} for details).
Each edge is depicted as a horizontal line connecting the genomic coordinates of the
parent and child nodes on the x-axis, and the y-axis shows the time of 
either the parent or the child, as chosen by the user.
The user can interact with the main ``genome browser'' window
using a set of controls on the top-right corner (provided by Bokeh),
allowing the user to pan and zoom as required. The histograms to the
right then summarise the edges as depicted in the browser window.
A similar browser interface is provided for 
nodes
(e.g. Fig~\ref{fig:Supplementary_Figure_3}) 
and mutations 
(e.g. Fig~\ref{fig:Supplementary_Figure_4}).
\texttt{Tsbrowse} also provides an interactive table viewer with flexible
search and sorting utilities, which is a valuable debugging utility for 
developers.

@@

\subsection{Applications}
The purpose of \texttt{tsbrowse} is to provide an interactive view of the 
\texttt{tskit} ARG data model to guide intuition, improve inference quality control
and facilitate debugging. An example of how we can deepen our understanding
using the genome-browser-like perspective provided by \texttt{tsbrowse}
is provided the simulated ARG of Fig~\ref{fig:Figure_2}, where 
the gap in the edges view corresponds to the characteristic dip in diversity
of a selective sweep. Comparing this ground truth to the ARGs inferred
by four different inference methods in Fig~\ref{fig:Supplementary_Figure_1},
we can see that there are substantial qualitative differences between the
results. These differences in ancestral haplotypes illustrated by 
the edges view are unlikely to be captured by the tree-by-tree distance
metrics usually used to evaluate
inference methods~\citep[e.g.][]{kelleher2019inferring,zhang2023biobank},
providing further motivation for new and improved ways to compare
simulated and inferred ARGs~\citep{fritze2024forest}.

Interest in ARG-based methods is burgeoning, but the methods are new 
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the 
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensible
element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
shows [a before and after view for 1000G data, where tsbrowse helped 
diagnose and resolve an opaque data problem].
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm. 
Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of 
ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear 
excess of long haplotypes in the very ancient past. These insights 
have helped guide development and 
may lead to significant improvements in performance.

@@

\section{Discussion} \label{sec:Discussion} 
Visualisation of tree topology is a central task in phylogenetics, 
and although numerous tools
exist~\citep[e.g.][]{huson2007dendroscope,vaughan2017icytree},
the methods can typically only handle a few hundred nodes.
Visualisation of large-scale tree topologies with millions of nodes
requires much more sophisticated approaches
to capture topological features at different scales, 
and is active research area~\citep{wong2022dynamic,kramer2023treenome}.
Adapting such methods, and integrating into \texttt{tsbrowse} to provide
a local tree viewer that operates at the million-node scale
is an important direction for future work.
An interactive viewer for the entire ARG topology, capturing 
semantic properties of the graph at a range of scales, is an even more 
ambitious goal, and would be a major asset for the field.

@@

\section*{Funding}
This work was supported by funding from the Biotechnology and Biological
Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
\textcolor{red}{add Novo funding and funding details for Duncan}. DMR is funded by 
a studentship from the Wellcome Programme in Genomic Medicine and Statistics.
JK acknowledges EPSRC (research grant EP/X024881/1),
NIH (research grants HG011395 and HG012473)
and the Robertson Foundation.

@@

\section{Supplementary Material} 
\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})} \label{subsec:sweep_simulation} 
Ancestral
histories of 300 samples were simulated with \texttt{SweepGenicSelection}
function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
in the recent past, a selective sweep was simulated with a beneficial allele
situated in the middle of a 5 Mb sequence. The frequency of the allele in the
population was set at 0.0001 at the beginning of the sweep. The allele fixed in
the population at a frequency of 0.9999. The strength of selection was set
using the selection coefficient, \textit{s = 0.25}. A time increment,
\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
the ARG at a rate of 1e-6. For simulating history before occurrence of the
sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
used until coalescence was achieved.

\subsection{Inference of selective sweep dataset (Suppl. Fig.
\ref{fig:Supplementary_Figure_1})} The following software were used to infer
ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
\texttt{Relate version 1.2.2} \cite{speidel2019method}, 
\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}. 
For all
methods, the following parameters of inference were used: \textit{recombination
rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
= 10,000}. Default values were used for all other inference parameters.

\subsection{Inference of 1000 Genomes dataset (Suppl. Fig.
\ref{fig:Supplementary_Figure_2}, Suppl. Fig.
\ref{fig:Supplementary_Figure_3})} The 1000 Genomes dataset was downloaded from
\textcolor{red}{[insert link]}. Ancestral fasta sequence for chromosome 20
(GRCh38) was downloaded from the Ensembl database. Inference was performed with
a Snakemake pipeline \textcolor{red}{(cite Ben's repo?)} using \texttt{tsinfer
version 0.3.3} for the long arm of chromosome 20 after filtering out duplicate
variant positions, variants with missing or low quality ancestral allele,
singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were included for
inference. For the inference in Suppl. Fig. 2A, all genomic regions were
included. For the inference in Suppl. Figs 2B and 3, genomic regions with a
site density lesser than 5 per kb were discarded. For Suppl. Fig. 2,
\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
with \textit{mutation rate = 1.29e-8}, setting all other parameters to default
values.

@@

\clearpage
\section{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A),
ARG-Needle (B), Relate (C) and SINGER (4) inferences of the truth dataset
simulated under a selective sweep model (in Figure 2 of the main text). For
SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
The X coordinate represents genomic position, each horizontal segment on the
plot shows the genomic coordinates that the edge spans, and Y coordinate shows
time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_1}
\end{figure}

\begin{figure}
    \centering
    % \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
    \caption{\textbf{A view of mutations in the 1000G WGS dataset.} A
screenshot of \texttt{tsbrowse}'s Mutations view for a X Mb region on
chromosome 17 inferred from from 3,202 participants from the 1000 Genomes Whole
Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. The genomic
position of the mutations is shown on the X axis, and the age of mutation is
represented on the Y axis. Panel B represents an inference after filtering out
low site-density regions in the samples.}
    \label{fig:Supplementary_Figure_2}
\end{figure}

>>>>

\section{Introduction}
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a
recombining
species, and are currently the subject of intense research
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
    wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
    speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
    gunnarsson2024scalable,deng2024robust} that practical application has
become possible. Varied applications have been recently proposed,
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
more powerful approaches to quantifying genetic
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree}.
While these developements are exciting, the performance 
of these new methods depends critically on the accuracy of the inferred ARGs.
Although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data is 
understudied. In particular, there is a critical lack of software 
infrastructure to support evaluation and quality control of 
inferred ARGs.

Visualisation is fundamentally important to data analysis.
Many specialised tools exist to aid the visual analysis and
quality control
of genome assembly~\citep{wick2015bandage,challis2020blobtoolkit},
read mapping~\citep{robinson2011integrative},
and variant
calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
for example.
At every stage of a bioinformatics pipeline, it is important to visualise
results to avoid artefacts and aid understanding of the data.
Genome browsers such as IGV~\citep{robinson2011integrative} and the
UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
data modalities, and are vital infrastructure for the field.
% . In the past few decades,
% genome browsers have become synonymous with sequence visualisation, by
% providing user-friendly, interactive interfaces to navigate genomic sequences
% and annotations. From early standalone tools to manage \textit{C. elegans} data
% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
% with the changing landscape of genomics. Modern genome browsers can load
% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
% The data security and scalability offered by this architecture ensures that
% genome browsers continue to be relevant in the biobank era.

@@

\subsection{Data model} \label{subsec:Data_Model} 
\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
ARGs~\citep{wong2024general}. This efficient ARG encoding 
is implemented by the 
\texttt{tskit} library~\citep{ralph2020efficiently}
and supported by most modern ARG
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
    haller2019tree,baumdicker2022efficient,
    adrion2020community,lauterbur2023expanding,korfmann2023weak,
    tsambos2023link,tagami2024tstrait},
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
    mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
In \texttt{tskit}, ARGs are encoded as a
collection of tables, storing information about the
nodes and edges that describe the graph topology and the sites
and mutations that encode the sequence variation (Fig.~\ref{fig:Figure_1}).
% For example, the Node table describes sampled or ancestral sequences
% and records their time of birth, and the individual and population they belong
% to. The Edge table describes the relationship between pairs of nodes, and
% records the node IDs and the genomic locations over which the edge spans. The
% Mutation table records mutation events in the history of the sample, including
% the change of allelic state, time of event and node in which the event
% occurred. 
% The Node and Edge tables describe the genealogical relationships
% between the sampled and ancestral sequences, 
% while the Mutation table describes
% sequence variation contained in the samples. 
% (see \hyperref[subsec:User_Interface]{User Interface}).

@@

The goal of \texttt{tsbrowse} is to provide interpretable
and interactive visual summaries of ARGs containing millions of nodes, edges
and mutations. Mutations, for example, have a clear interpretation
when plotted on genome coordinate vs time axes, but at this scale
the data density is far too high to simply plot each mutation as a point.
We overcome this problem by using Datashader and the wider
Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
at the requested resolution on the server, sends the image to the web-browser
for display, and dynamically updates as the user interactively navigates the
ARG.
This approach allows us to summarise very large ARGs interactively;
Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of
\texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
ARG with around 2.7 million nodes and edges.

@@

\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
    \centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{Screenshot of the Edges tab in \texttt{tsbrowse}.
        Visualisation of the edges in a simulated ARG with a strong selective
        sweep in the middle of the genome (see Supplement for details).
        Edges are shown in the main browser pane on the left, with additional
        histograms summarising the edges on the right.
        The effect of the sweep can be seen by the lack of edges crossing the
        centre of the simulated chromosome due to an excess of recent coalescent
        events. Moving away from the focal site, the oldest edges are the
        first to rejoin the ARG, followed by more recent edges, resulting in a
        wedge-like pattern of missing edges in the centre.
        \label{fig:Figure_2}}
\end{figure}
% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
% visualisation pipelines even for large datasets. A key challenge with
% large-scale data visualisation is to maintain a responsive user interface.
% \texttt{Datashader} aggregates data and renders only the necessary data points
% at a given zoom level, thereby minimising the computational load on the
% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
% with individual data points, allowing users to gain deeper insights from the
% data. The \texttt{serve} step accepts an optional input file containing
% sequence annotations. This allows users to overlay contextual information about
% genes or other sequence features. 
The user interface is presented as a dashboard, with pages to
describe various aspects of the ARG.
Figure~\ref{fig:Figure_2}
demonstrates the user interface using the Edges page as an example.
The plot on the left is an overview of the 33,929 edges
in a simulated ARG with a strong selective sweep (see
Sec.~\ref{subsec:sweep_simulation} for details).
Each edge is depicted as a horizontal line connecting the genomic coordinates
of the
parent and child nodes on the x-axis, and the y-axis shows the time of
either the parent or the child, as chosen by the user.
The user can interact with the main ``genome browser'' window
using a set of controls on the top-right corner (provided by Bokeh),
allowing the user to pan and zoom as required. The histograms to the
right then summarise the edges as depicted in the browser window.
A similar browser interface is provided for
nodes
(e.g. Fig~\ref{fig:Supplementary_Figure_3})
and mutations
(e.g. Fig~\ref{fig:Supplementary_Figure_4}).
\texttt{Tsbrowse} also provides an interactive table viewer with flexible
search and sorting utilities, which is a valuable debugging utility for
developers.

@@

\subsection{Applications}
The purpose of \texttt{tsbrowse} is to provide an interactive view of the
\texttt{tskit} ARG data model to guide intuition, improve inference quality
control and facilitate debugging. An example of how we can deepen our understanding
using the genome-browser-like perspective provided by \texttt{tsbrowse}
is given in the simulated ARG of Fig~\ref{fig:Figure_2}, where
the gap in the edges view corresponds to the characteristic dip in diversity
of a selective sweep. Comparing this ground truth to the ARGs inferred
by four different inference methods in Fig~\ref{fig:Supplementary_Figure_1},
we can see that there are substantial qualitative differences between the
results. These differences in ancestral haplotypes illustrated by
the edges view are unlikely to be captured by the tree-by-tree distance
metrics usually used to evaluate
inference methods~\citep[e.g.][]{kelleher2019inferring,zhang2023biobank},
providing further motivation for new and improved ways to compare
simulated and inferred ARGs~\citep{fritze2024forest}.

Interest in ARG-based methods is burgeoning, but the methods are new
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensable
element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
shows a region of the 1000G data with gaps in site density that are spanned by
 exceptionally long edges, which are likely to bias downstream statistics.
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm.
Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of
ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear
excess of long haplotypes in the very ancient past. These insights
have helped guide development and
may lead to significant improvements in performance.

@@

\section{Discussion} \label{sec:Discussion}
Visualisation of tree topology is a central task in phylogenetics,
and although numerous tools
exist~\citep[e.g.][]{huson2007dendroscope,vaughan2017icytree},
the methods can typically only handle a few hundred nodes.
Visualisation of large-scale tree topologies with millions of nodes
requires much more sophisticated approaches
to capture topological features at different scales,
and is an active research area~\citep{wong2022dynamic,kramer2023treenome}.
Adapting such methods, and integrating into \texttt{tsbrowse} to provide
a local tree viewer that operates at the million-node scale
is an important direction for future work.
An interactive viewer for the entire ARG topology, capturing
semantic properties of the graph at a range of scales, is an even more
ambitious goal, and would be a major asset for the field.

@@

\section*{Funding}
This work was supported by funding from the Biotechnology and Biological
Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
\textcolor{red}{add Novo funding}. DMR is funded by 
a studentship from the Wellcome Programme in Genomic Medicine and Statistics.
JK acknowledges EPSRC (research grant EP/X024881/1),
NIH (research grants HG011395 and HG012473)
and the Robertson Foundation.

@@

\section{Supplementary Material}
\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
\label{subsec:sweep_simulation}
Ancestral
histories of 300 samples were simulated with \texttt{SweepGenicSelection}
function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
in the recent past, a selective sweep was simulated with a beneficial allele
situated in the middle of a 5 Mb sequence. The frequency of the allele in the
population was set at 0.0001 at the beginning of the sweep. The allele fixed in
the population at a frequency of 0.9999. The strength of selection was set
using the selection coefficient, \textit{s = 0.25}. A time increment,
\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 1e-8 per base pair per generation. For simulating history before occurrence of the
sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
used until coalescence was achieved.

\subsection{Inference of selective sweep dataset (Suppl. Fig.
    \ref{fig:Supplementary_Figure_1})} The following software were used to infer
ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
\texttt{Relate version 1.2.2} \citep{speidel2019method}, 
\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}.
For all
methods, the following parameters of inference were used: \textit{recombination
    rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
    = 10,000}. Default values were used for all other inference parameters.

\subsection{Inference of 1000 Genomes dataset (Suppl. Fig.
\ref{fig:Supplementary_Figure_2}, Suppl. Fig.
\ref{fig:Supplementary_Figure_3})} The 1000 Genomes dataset was downloaded from
\texttt{\url{https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/}}.
Ancestral fasta sequences for chromosomes 17 and 20 (GRCh38) were downloaded
from the Ensembl database. Inference was performed with a Snakemake pipeline
(\texttt{\url{https://github.com/benjeffery/tsinfer-snakemake/}}) using
\texttt{tsinfer version 0.3.3} for the long arms of chromosome 17 and 20 after
filtering out duplicate variant positions, variants with missing or low quality
ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
included for inference. For the inference in Suppl. Fig.
\ref{fig:Supplementary_Figure_2}, all genomic regions were included. For the
inference in Suppl. Fig. \ref{fig:Supplementary_Figure_3}, genomic regions with
a site density lesser than 5 per kb were discarded. For Suppl. Fig.
\ref{fig:Supplementary_Figure_2}, \texttt{tsdate version 0.2.1} was used to
estimate the age of ancestral nodes with \textit{mutation rate = 1.29e-8},
setting all other parameters to default values.

@@

\clearpage
\section{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
ARG-Needle, Relate and SINGER inferences of the truth dataset
simulated under a selective sweep model (in Figure 2 of the main text). For
SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
The X coordinate represents genomic position, each horizontal segment on the
plot shows the genomic coordinates that the edge spans, and Y coordinate shows
time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}} 
Screenshots of \texttt{tsbrowse}'s Mutations view (A) and Edges view (B) for a
 600 kB region of chromosome 17 inferred from 3,202 participants from the 1000 
 Genomes Whole Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. 
 The poor performance of \texttt{tsinfer} in this variant-poor region is evidenced
  by the long edges spanning gaps in mutation density.}
    \label{fig:Supplementary_Figure_2}
\end{figure}
